### 2021-02-22 paper.Rmd
more intro

Data commonly arrives with more than two measured variables, which makes it more complicated to plot on a page. With multiple variables, especially if there is some association between variables, this would be called multivariate or high-dimensional data. When the variables are all numeric, or quantitative, visualisation often relies on some form of dimension reduction. This can be done by taking linear projections, for example, principal component analysis [@Hotelling1933-of] or linear discriminant analysis [@Fisher1936], or nonlinear if one considers techniques like multidimensional scaling (MDS) [@Kruskal1964-do] or t-Distributed Stochastic Neighbour Embedding (t-SNE) [@Van_der_Maaten2008-qa]. For the purposes here, the focus is on linear projections, in particular as provided by the grand tour [@Asimov1985-xr; @Buja2005-cx]. Interactive and dynamic displays can provide information beyond what can be achieved in a static display. The grand tour shows a smooth sequence of interpolated low dimensional projections, and allows the viewer to extrapolate from the low-dimensional shapes to the multidimensional distribution. It is particularly useful for detecting clusters, outliers and non-linear dependence.

>>>>

Data commonly arrives with more than two measured variables, which makes it more complicated to plot on a page. With multiple variables, especially if there is some association between variables, this would be called multivariate or high-dimensional data. When the variables are all numeric, or quantitative, visualisation often relies on some form of dimension reduction. This can be done by taking linear projections, for example, principal component analysis [@Hotelling1933-of] or linear discriminant analysis [@Fisher1936]. It is also common to reduce dimension nonlinearly with techniques like multidimensional scaling (MDS) [@Kruskal1964-do] or t-Distributed Stochastic Neighbour Embedding (t-SNE) [@Van_der_Maaten2008-qa].

The area of visualising high-dimensions is very large, so the focus of this review will be on linear projections, in particular as provided by the grand tour [@Asimov1985-xr; @Buja2005-cx]. An overview of the technique, and recent developments is provided, along with how these techniques can be used in a variety of applications, and to help understand nonlinear dimension reduction results.

A tour can be considered to be a dynamic graphic, because it shows a smooth sequence of projections over time, ideally with controls that allow stopping, reversing, changing direction, or going forward again. It can be useful to embed a tour into an interactive graphics system, where plots can be queried and elements highlighted (see for e.g. @Swayne2003-cc or @Tierney1991). To create the smooth sequence, a geodesic interpolation is computed between consecutive frames.  It allows the viewer to extrapolate from the low-dimensional to  shapes corresponding to multivariate distribution, and is particularly useful for detecting clusters, outliers and non-linear dependence.
